Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-29 draft.tex
Revised Version from Anja

% \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
\abstract{ 
The goal of epidemiology is to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess the predictive quality of image-derived variables for lower back pain.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as the spine canal course, its curvature and torsion.
%%
We analyze the predictive power of shape-describing variables for non-image variables, e.g., pain indicators.
%%
Therefore, we automatically extract a decision tree classifier for each non-image variable.
%%
Furthermore, we provide an IVA technique to compare the classifiers with interactive scatterplot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors.
%%
A correlation between lumbar spine shape and lower back pain disease was not existent in the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, correlations between non-image variables were successfully detected and characterized.
}

@@

%For epidemiological research, cohort studies are acquired.
Cohort studies are an instrument of epidemiological research.
%For epidemiological research, cohort studies are carried out.
%%
Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
%%
Back pain is one of the most frequent diseases in the western civilization and gains importance due to the population aging.
%%
Epidemiological analysis of back pain aims to characterize the disease by determining risk factors.
%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, an automatic classification approach for characterization of back pain based on lumbar back spine attributes is still missing.
%%
Furthermore, previous analyses mainly focused on a single variable (e.g. \emph{BMI}) or a small subset of variables.
%%

We present a thorough analysis of the image-derived data of a cohort study including lumbar spine datasets.
%%
We analyze their discriminative power for different back pain characteristics.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis,
%%
\item The detection of correlations between image-based and socio-demographic as well as medical variables for an improved hypothesis generation,
%%
\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
%%
\end{itemize}
%%
The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To prove a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
For example, the effect of nutrition on prostate cancer requires age-normalization due to the increased incidence for older subjects.
%%
%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
%\subsection{Epidemiological Data}
\subsection{Challenges of Epidemiological Data Analysis}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies analyzing one specific disease and its influence on the human body.
%%
We focus on large scale cohort study data.
%%
These studies aim at collecting as much data as possible for each subject.
%%
These data can be analyzed regarding many diseases and conditions.
%%

@@

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
%The lower (\emph{lumbar}) spine is the most stressed spine part .
%%
Lower back pain is one of the most frequent diseases in the western civilization \cite{Hoy2010}.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, physical activity and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are unknown.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

@@

Zhang et al. analyze subject groups in a web-based linked view system \cite{Zhang}.

@@

@@

\paragraph{The Error Term. }
Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {n}$ where $n$ is the number of subjects.
%%
The metric usually works well for variables with uniform distribution.
%%
It distorts the result for other distribution types.
%%
For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
%%
Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
%%
%Therefore the discriminative power of each dimension in the equation
%error = 1 - \frac{\[ \sum_{m=0}^M \] \frac{M_{classifiedCorrectly}}{M_all}}{number of dimensions}
\begin{equation}
error = 1 - \frac{\sum_{m=0}^M \frac{correctlyClassified_{m}}{m}}{number_{dimensions}}
\end{equation}
$M$ represents the set of manifestation of each variable.
%%
The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
% The error rate has to be judged depending on the number of classes of the target variable.
% %%
% An error rate of 50\% for a target variable with two manifestations makes the classifier equal to throwing a coin.
% %%
% The same error rate is acceptable a target variable with more than two manifestations.
% %%
% Picking a value for a variable of four manifestations with an error rate of 50\% is better than the random selection (error rate of 75\%).
% %%
% The following scales the error rate so that 0 denotes perfect classification and 1 is equal to random selection.
% %%
% \begin{equation}
% error_{normalized} = \frac{error}{100 - \frac{100}{class number}}
% \end{equation}
%%
%%
We consider a result rated below $0.25$ a good classification.
%%
It allows for comparability of error rates between variables with different manifestation count.

@@

%\com{Potentially skip these due to the confounding of body size}
\subsection{Body-Mass-Index Groups}
%%
\emph{Gender} could be classified for each \emph{BMI} group using \emph{mean curvature} and \emph{curvature angle}.
%%
The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
%%
The starting age of smoking could be characterized well with an error between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with \emph{BMI} of \emph{$30.5-48$}.
%%
The result however is probably overfitted to the data due to tree sizes between $14$ and $16$.
%%

@@

\subsection{Size Groups}
%%
Many previously described results are confounded by \emph{subject size}.
%%
Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine variability variable is enough for predicting other variables or not.
%%
Dividing subjects into \emph{body size} groups potentially highlights classifications not confounded by \emph{body size}.
%%
\paragraph{Large Decision Trees. }
\emph{Back pain}-associated variables can be explained for various \emph{size}-groups, but we could not extract universal rules.
%%
\emph{Radiating back pain} could be described with error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with an error between $0.42$ to $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.

@@

Another observation during the analysis is that large decision trees achieve low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\paragraph{Applicability.}
%%
Classification methods based on decision trees have proven to be useful for assessing the discriminative power of a variable set.
%%
Their ability to consider variable combination makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, the results are harder to assess and to abstract.
%%
Our method to plot derived metrics and custom-tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our interactive visualization.
%%
Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%

@@

%In our future work, we will focus on more precise models for extracting measures described in Section~\ref{sec:ProposedFeatures}.
In our future work, we will focus on more precise models for extracting measures.
%%
Dented vertebras are a early sign of a pathological deformation, so we want to capture this by segmenting the top and bottom point of each vertebra center.
%%
Another interesting metric is the spine canal thickness over the whole spine, capturing early signs of herniated disc disease.
%%
As another focus, we want to include the method into existing visual analytics methods designed for analyzing shape information for epidemiological data \cite{Klemm2014VIS}.
%%
\\\\
Combining the power of statistical analysis, visual analytics and data mining techniques is essential for analyzing increasingly complex heterogenous population data.
%%
These methods do not aim to replace the traditional epidemiological workflow, but rather intervenes at the weak points of standard statistical methods.

>>>>

% \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
\abstract{ 
The goal of epidemiology is to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess the predictive quality of image-derived variables for lower back pain.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as the spine canal course, its curvature and torsion.
%%
We analyze the predictive power of shape-describing variables for non-image variables, e.g. pain indicators.
%%
Therefore, we automatically extract a decision tree classifier for each non-image variable.
%%
Furthermore, we provide an IVA technique to compare the classifiers with interactive scatterplot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors.
%%
A correlation between lumbar spine shape and lower back pain disease was not existent in the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, correlations between non-image variables were successfully detected and characterized.
}

@@

%For epidemiological research, cohort studies are acquired.
Cohort studies are an instrument of epidemiological research.
%For epidemiological research, cohort studies are carried out.
%%
Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
%%
Back pain is one of the most frequent diseases in the western civilization and gains importance due to the population aging \cite{Hoy2010}.
%%
Epidemiological analysis of back pain aims to characterize the disease by determining risk factors.
%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, an automatic classification approach for characterization of back pain based on lumbar back spine attributes is still missing.
%%
Furthermore, previous analyses mainly focused on a single variable (e.g. \emph{BMI}) or a small subset of variables.
%%

We present a thorough analysis of the image-derived data of a cohort study including lumbar spine datasets.
%%
We analyze their discriminative power for different back pain characteristics.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis,
%%
\item The detection of correlations between image-based and socio-demographic as well as medical variables for an improved hypothesis generation,
%%
\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
%%
\end{itemize}
%%
The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age \cite{Fletcher}.
%%
For example, the effect of nutrition on prostate cancer requires age-normalization due to the increased incidence for older subjects.
%%
%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
%\subsection{Epidemiological Data}
\subsection{Challenges of Epidemiological Data Analysis}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies analyzing one specific disease and its influence on the human body.
%%
We focus on large scale cohort study data.
%%
These studies aim at collecting as much data as possible for each subject.
%%
The collected data can be analyzed regarding many diseases and conditions.
%%

@@

Modern cohort studies often comprise medical image data.
%%
The analysis of these data is challenging.
%These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
%The lower (\emph{lumbar}) spine is the most stressed spine part .
%%
Lower back pain is one of the most frequent diseases in the western civilization \cite{Hoy2010}.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, physical activity and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are unknown.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

@@

\cite{Zhang} analyze subject groups in a web-based linked view system.

@@

%	Women have a higher live expectance than men, hence their higher share in the old age group.
%	Also, women are on average smaller than men, hence the larger shape similarity with older subjects.

@@

\paragraph{The Error Term. }
Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {total_{n}}$ where $n$ is the number of subjects.
%%
The metric usually works well for variables with uniform distribution.
%%
It distorts the result for other distribution types.
%%
For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
%%
Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
%%
%Therefore the discriminative power of each dimension in the equation
%error = 1 - \frac{\[ \sum_{m=0}^M \] \frac{M_{classifiedCorrectly}}{M_all}}{number of dimensions}
\begin{equation}
error = 1 - \frac{\sum_{m=0}^M \frac{correctlyClassified_{m}}{total_{m}}}{number_{dimensions}}
\end{equation}
$M$ represents the set of manifestation of each variable.
%%
The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
% The error rate has to be judged depending on the number of classes of the target variable.
% %%
% An error rate of 50\% for a target variable with two manifestations makes the classifier equal to throwing a coin.
% %%
% The same error rate is acceptable a target variable with more than two manifestations.
% %%
% Picking a value for a variable of four manifestations with an error rate of 50\% is better than the random selection (error rate of 75\%).
% %%
% The following scales the error rate so that 0 denotes perfect classification and 1 is equal to random selection.
% %%
% \begin{equation}
% error_{normalized} = \frac{error}{100 - \frac{100}{class number}}
% \end{equation}
%%
%%
%We consider a result rated below $0.5$ an result of interest, a value below $0.25$ represents a good classification.
We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
%%
It allows for comparability of error rates between variables with different manifestation count.

@@

%\com{Potentially skip these due to the confounding of body size}
\subsection{Body-Mass-Index Groups}
%%
\emph{Gender} could be classified for each \emph{BMI} group using \emph{mean curvature} and \emph{curvature angle}.
%%
The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}. 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
%%
The starting age of smoking could be characterized well with an error between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with \emph{BMI} of \emph{$30.5-48$}.
%%
The result however is probably overfitted to the data due to tree sizes between $14$ and $16$.
%%

@@

\subsection{Size Groups}
%%
Many previously described results are confounded by \emph{subject size}.
%%
Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine variability parameter is enough for predicting other variables or not.
%%
Dividing subjects into \emph{body size} groups potentially highlights classifications not confounded by \emph{body size}.
%%
\paragraph{Large Decision Trees. }
\emph{Back pain}-associated variables can be explained for various \emph{size}-groups, but we could not extract universal rules.
%%
\emph{Radiating back pain} could be described with error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with an error between $0.42$ to $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.

@@

Another observation during the analysis is that large decision trees achieve low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\paragraph{Applicability.}
%%
Classification methods based on decision trees have proven to be useful for assessing the discriminative power of a variable set.
%%
Their ability to consider variable combination makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, the results are more challenging to assess and to abstract.
%%
Our method to plot derived metrics and custom-tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our interactive visualization.
%%
Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%

@@

%In our future work, we will focus on more precise models for extracting measures described in Section~\ref{sec:ProposedFeatures}.
In our future work, we will focus on more precise models for extracting measures.
%%
Dented vertebras are a early sign of a pathological deformation, so we want to capture this by segmenting the top and bottom point of each vertebra center.
%%
Another interesting metric is the spine canal thickness over the whole spine, capturing early signs of herniated disc disease.
%%
As another focus, we want to include the method into existing visual analytics methods designed for analyzing shape information for epidemiological data \cite{Klemm2014VIS}.
%%
\\\\
Combining the power of statistical analysis, visual analytics and data mining techniques is essential for analyzing increasingly complex heterogenous population data.
%%
These methods do not aim to replace the traditional epidemiological workflow, but rather intervenes at the weak points of standard statistical methods.
%%
Our method provides a novel way to gain insight into these complex data sets and amplifies hypotheses generation.
